---
title: "Week 3: Getting Data"
subtitle: "CDAE 7990 · Applied Data Science & Visualization"
author: "Andrew Van Leuven"
date: "September 17, 2026"
format:
  revealjs:
    theme: [default, ../slides.scss]
    slide-number: true
    show-slide-number: speaker
    footer: "CDAE 7990 · Week 3"
    logo: ""
    transition: fade
    code-line-numbers: false
    scrollable: true
    date-format: long
execute:
  echo: true
  eval: true
---


```{r}
#| include: false
library(tidyverse)
options(pillar.print_max = 5, pillar.print_min = 5, width = 72)
```

## Welcome to Morrill Hall 005

- Our new room for the rest of the semester: a basement computer lab, so
  there's a machine at every seat
- Tonight is a tour of where data comes from, beyond the file you already
  have sitting in your project
- We'll land on exactly what HW2's join step needs

## Today's Plan

- **Housekeeping**: nothing new is due this week
- **Reading files, revisited**: `skip`, `na`, and messy real-world exports
- **A quick tour of `tidyr`**: wide data vs. long data
- **APIs**: asking a website for data instead of clicking through pages
- **Web scraping**: a last resort, and why
- *Break*
- **Joins**: combining two data frames, live, using HW2's own data

# Housekeeping {background-color="#007155"}

## Nothing New Due This Week

- HW2 (due Sept. 27) is still your only open assignment
- Tonight's material, especially joins, is exactly what HW2's step 5 needs
- We'll do that join together before you leave, using your own files

# Getting Data {background-color="#007155"}

## Where Data Comes From

- A file someone hands you, or you download: CSV, Excel, and the like
- A live API you query with code
- A web page you scrape, when nothing else is available
- R4DS Ch. 7-8 covers the first in depth. APIs and scraping aren't in R4DS
  at all, that material is further along in the book than we've gotten

## Reading Files, Revisited {.smaller}

```{r}
#| eval: false
read_csv(
  "data/hw2a.csv",
  skip = 3,
  na = c("", "NA", "(NA)")
)
```

- You've used `read_csv()` since HW2's starter code
- `skip` jumps past header rows a file doesn't need, common in government
  exports
- `na` tells readr which strings actually mean "missing." BEA uses `(NA)`,
  not R's default `NA`
- readr prints a column specification when it loads. Always glance at it. A
  date or number read in as text usually means a stray character snuck into
  that column

## Wide vs. Long Data {.smaller}

```{r}
bea_wide <- tibble(
  county_name = c("Addison, VT", "Chittenden, VT", "Rutland, VT"),
  pop_2020 = c(37363, 168323, 60572),
  pop_2021 = c(37162, 169660, 60381),
  pop_2022 = c(36777, 170533, 59949)
)

bea_wide
```

- One row per county, but each year gets its own column
- This is what you get if you ever download **multiple years** from BEA at
  once, instead of the single year HW2 asked for

## `pivot_longer()`: Wide to Long {.smaller}

```{r}
bea_long <- bea_wide |>
  pivot_longer(
    cols = starts_with("pop_"),
    names_to = "year",
    names_prefix = "pop_",
    values_to = "population"
  )

bea_long
```

- One row per county *and* year, instead of one row per county
- Almost everything in dplyr and ggplot2 wants data in this shape
- `names_prefix` strips `"pop_"` so the year column just holds the number

## `pivot_wider()`: The Reverse {.smaller}

```{r}
bea_long |>
  pivot_wider(names_from = year, values_from = population)
```

- Takes values from one column and spreads them into new columns, one per
  unique value
- Good for a summary table you want to read by eye, less useful once you
  need to filter or group by year again

# Getting Data from an API {background-color="#007155"}

## What's an API?

- A structured way to ask a website for data and get back something other
  than a webpage, usually JSON
- You build a URL with your request baked in, and the server sends back
  data instead of a rendered page
- Census, BLS, and BEA all publish APIs. Some want a free registration key,
  a few don't

## A Live Example: the Census API {.smaller}

```{r}
#| eval: false
library(jsonlite)

url <- "https://api.census.gov/data/2020/dec/pl?get=NAME,P1_001N&for=state:*"
census_raw <- fromJSON(url)

head(census_raw)
```

- `P1_001N` is the Census's code for total population
- No key needed for a request this small, larger ones want one, it's free
- What comes back is a plain character matrix, not a tidy tibble yet

## Turning JSON into a Tidy Data Frame {.smaller}

```{r}
#| eval: false
census_pop <- census_raw |>
  as_tibble(.name_repair = "minimal") |>
  set_names(census_raw[1, ]) |>
  slice(-1) |>
  mutate(P1_001N = as.numeric(P1_001N))
```

- The first row of the matrix holds the column names, not data. `set_names()`
  promotes it, `slice(-1)` drops it
- This cleanup is the same shape every time: fetch, then reshape
- Week 8's `tidycensus` package does this exact cleanup for you in one
  function call. Tonight you're seeing what it's hiding

## Web Scraping: A Quick Look {.smaller}

```{r}
#| eval: false
library(rvest)

page <- read_html("https://en.wikipedia.org/wiki/List_of_counties_in_Vermont")
counties <- page |>
  html_element("table.wikitable") |>
  html_table()
```

- Some data only exists inside a web page, with no file or API behind it
- `rvest` reads a page's HTML and pulls out a specific table or element
- Fragile: if the page's layout changes, your code breaks with no warning
- Use it last, after checking for a file or an API. If it's an NBRC page,
  ask first rather than scraping on their behalf

## Break? {background-color="#007155"}

# Joins: Combining Two Data Frames {background-color="#007155"}

## Why Joins

- HW2 step 5 asks you to combine `df` and `df2` using a shared column,
  `county_fips`
- A join lines up rows from two tables using one or more shared columns,
  called **keys**
- Nothing here is BEA-specific. Joins work the same way on any two data
  frames that share a key

## A Small Example {.smaller}

```{r}
pop <- tibble(
  county_fips = c("50001", "50003", "50007"),
  county_name = c("Addison", "Bennington", "Chittenden"),
  population = c(37363, 34478, 168323)
)

income <- tibble(
  county_fips = c("50001", "50003", "50011"),
  income = c(58000, 52000, 71000)
)
```

- Both tables share `county_fips`, the key
- The counties don't fully overlap: 50007 is only in `pop`, 50011 only in
  `income`

## `left_join()`: Keep Everything on the Left {.smaller}

```{r}
pop |> left_join(income, by = "county_fips")
```

- Every row of `pop` survives, matched or not
- Chittenden (50007) has no match in `income`, so it gets `NA`
- 50011's income is dropped. It wasn't in `pop` to begin with

## Other Join Types, Briefly

| Join | Keeps |
|---|---|
| `left_join()` | Everything on the left, matched where possible |
| `inner_join()` | Only rows that match in both |
| `full_join()` | Everything from both, matched where possible |

`left_join()` is the one you'll reach for almost every time: you have a main
table and you're adding to it.

## Watch Out: Duplicate Column Names {.smaller}

- If both tables have a column with the same name that isn't the key (both
  have `county_name`, say), the join keeps both and adds `.x` and `.y`
  suffixes
- HW2 warns you about exactly this. Two fixes:
  - `select()` away the extra column from one side before joining
  - Join on every shared column: `by = c("county_fips", "county_name", "state")`

## Doing HW2 Step 5, Live {.smaller}

```{r}
#| eval: false
df |>
  select(county_fips, population_2024) |>
  left_join(df2, by = "county_fips")
```

- Trimming `df` down to just the key and the one column you need sidesteps
  the `.x`/`.y` problem entirely
- Swap in your own file and column names. Everyone's `df2` will look
  different
- Once this runs, HW2 step 6 just asks you to describe what the join added

## Your Turn (15 min) {.smaller}

Using `pop` and `income` from tonight, or your own HW2 `df`/`df2`:

1. Try `inner_join()` instead of `left_join()` on `pop` and `income`. How
   many rows come back, and why?
2. Pivot `bea_wide` to long format, then find the county with the biggest
   year-over-year change
3. Run your actual HW2 join. If you get a `.x`/`.y`, fix it one of the two
   ways from tonight's slide

Work with a neighbor. Reading the whole error message aloud still works.

## Your Turn: One Way to Do It {.smaller}

```{r}
#| eval: false
# 1
pop |> inner_join(income, by = "county_fips")
# Only 2 rows: just the counties present in both tables

# 2
bea_long |>
  group_by(county_name) |>
  mutate(change = population - lag(population)) |>
  filter(!is.na(change))

# 3
df |>
  select(county_fips, population_2024) |>
  left_join(df2, by = "county_fips")
```

## Before Next Class

- [ ] Nothing new due this week, but get HW2 (due Sept. 27) working if
      tonight's join didn't click. Office hours or email, any time
- [ ] Read R4DS [Ch. 9 (Layers)](https://r4ds.hadley.nz/layers) and
      [Ch. 10 (Exploratory data analysis)](https://r4ds.hadley.nz/eda), and
      skim [Ch. 11 (Communication)](https://r4ds.hadley.nz/communication)

Next week: exploratory data analysis and your first `ggplot2` charts.
