---
title: "Week 2: Into the Tidyverse"
subtitle: "CDAE 7990 · Applied Data Science & Visualization"
author: "Andrew Van Leuven"
date: "September 10, 2026"
format:
  revealjs:
    theme: [default, ../slides.scss]
    slide-number: true
    show-slide-number: speaker
    footer: "CDAE 7990 · Week 2"
    logo: ""
    transition: fade
    code-line-numbers: false
    scrollable: true
    date-format: long
execute:
  echo: true
  eval: true
---


```{r}
#| include: false
library(tidyverse)
options(pillar.print_max = 5, pillar.print_min = 5, width = 72)
```

## Welcome to Lafayette L202

- Our first in-person session. Let's do a quick re-introduction (name, day job, hometown).
- Today is hands-on. Open your Posit Cloud project and keep it open all night.
- Tonight's goal is to get comfortable with R and use it to manipulate some real-world data

## Today's Plan

- **Housekeeping**: HW1 is due Sunday, and HW2 needs to get into your project
- **R warm-up**: objects, functions, packages, and where things live in RStudio
- **dplyr, part 1**: working with rows and columns
- *Break*
- **dplyr, part 2**: groups and summaries
- **Code style**, and a preview of HW2

# Housekeeping {background-color="#007155"}

## HW1: Due Sunday (Sept. 13) by 11:59pm

Two notes on the version in your project:

- **Make your two vectors the same length.** Step 4 combines them into a data frame. If one has 5 items and the other has 6, R will throw an error.
- **Click Render before you save.** Confirm it knits to HTML with no errors. I grade by opening your copy, so an error means I see an empty page.

## Get HW2 Into Your Project

The `hw2.qmd` in your project is an old placeholder. Let's swap in the real one together.

1. Click the **Terminal** tab, next to Console (not the Console itself!)
2. Copy and paste these two lines, then press Enter:

```bash
git fetch
git checkout origin/main -- homework/hw2.qmd homework/data README.md
```

3. Open `homework/hw2.qmd`. It should start with instructions for downloading BEA data, and you should see a new `homework/data` folder.

## Get HW2 Into Your Project: What to Expect

- **No output is normal.** If nothing prints, it worked.
- The Git pane may list a few files as changed. Ignore it.
- **Your HW1 work is not touched**
- Please **don't** use `git pull` or the Pull button. From here on, every new homework comes out with these same two lines and a different file name.

# R Warm-Up {background-color="#007155"}

## Where Things Live in RStudio

| Pane | What it's for |
|---|---|
| **Editor** (top left) | Your `.qmd` files. This is where your work lives |
| **Console** (bottom left) | Runs R right now. Nothing typed here is saved |
| **Terminal** (tab next to Console) | Talks to the computer, not R. We only use it for `git` |
| **Environment** (top right) | Every object you've created this session |
| **Files** (bottom right) | Your project's folders. Upload data here |

## Working in a Quarto Document

- Code goes in **chunks**, gray boxes that start with ```` ```{r} ````
- Run one line: **Cmd/Ctrl + Enter**
- Run a whole chunk: **Cmd/Ctrl + Shift + Enter**, or the green ▶ arrow
- **Render** runs *everything* from top to bottom in a fresh session, and builds the HTML file
- If it works when you run it by hand but fails at Render, you probably created something in the Console that isn't in the document

## Objects and Assignment

```{r}
towns <- c("Burlington", "Montpelier", "St. Albans")
pop <- c(44743, 8074, 6876)

pop
```

- `<-` means "store this under that name"
- `c()` *combines* values into a **vector**
- A name like `pop` doesn't show anything until you run it

## Functions

```{r}
mean(pop)
max(pop)
round(mean(pop), digits = 0)
```

- A function takes **arguments** in parentheses and returns a result
- Functions can nest. R works from the inside out.
- Stuck? Type `?round` in the Console to see its help page.

## Data Frames

```{r}
vt <- data.frame(town = towns, population = pop)
vt
```

- A data frame is a set of vectors of the **same length**, side by side
- Rows are observations, columns are variables
- This is the only kind of object we'll use for most of the semester

## Packages

- Base R can do a lot, but packages add new functions
- **Install once** per project: `install.packages("tidyverse")`
  - Already done for you in Posit Cloud
- **Load every session**, at the top of every document:

```{r}
#| eval: false
library(tidyverse)
```

- If R says it `could not find function`, you probably forgot to load the package

## The Tidyverse

A family of packages that share a design philosophy, and that are built to work together.

| Package | Job |
|---|---|
| **readr** | Read data in (`read_csv()`) |
| **dplyr** | Transform it (tonight's topic) |
| **tidyr** | Reshape it |
| **stringr** | Work with text |
| **ggplot2** | Visualize it (Weeks 4–5) |

## Tonight's Data: `midwest` {.smaller}

```{r}
glimpse(midwest)
```

## Tonight's Data: `midwest`

- 437 counties in IL, IN, MI, OH and WI, from the 2000 Census
- Comes with the tidyverse (inside ggplot2), so there's nothing to download
- It's the same shape as your HW2 data: **one row per county**
- Useful columns: `county`, `state`, `poptotal`, `percollege`, `percbelowpoverty`, and `inmetro` (1 = metro county, 0 = non-metro)

## Break? {background-color="#007155"}

# dplyr, Part 1: Rows and Columns {background-color="#007155"}

## How Every dplyr Verb Works

- The first argument is always a data frame
- The other arguments say what to do with it, using column names **without quotes**
- The result is always a **new** data frame
- The original is left unchanged. To keep a result, assign it with `<-`.

## The Pipe: `|>` {.smaller}

Read `|>` as **"and then."**

```{r}
#| eval: false
# nested: read inside out
arrange(filter(midwest, state == "WI"), poptotal)

# piped: read top to bottom
midwest |>
  filter(state == "WI") |>
  arrange(poptotal)
```

To get `|>` instead of `%>%` (in RStudio on your machine), go to Tools → Global Options → Code and check *Use native pipe operator*. This is already enabled in Posit Cloud!

## `filter()`: Keep Rows That Match {.smaller}

```{r}
midwest |>
  filter(state == "WI", inmetro == 0) |>
  select(county, state, poptotal, inmetro)
```

A comma between conditions means **and**.

## `filter()`: Comparisons {.smaller}

| Operator | Meaning |
|---|---|
| `==` | equals (two `=` signs!) |
| `!=` | does not equal |
| `>`, `>=`, `<`, `<=` | greater/less than |
| `&`, `|` | and, or |
| `%in%` | matches any value in a list |

```{r}
#| eval: false
midwest |> filter(state == "MI" | state == "OH")
midwest |> filter(state %in% c("MI", "OH"))   # same thing, less typing
```

## `filter()`: Mistakes Everyone Makes {.smaller}

```{r}
#| eval: false
midwest |> filter(state = "WI")          # one = sign
midwest |> filter(state == WI)           # forgot the quotes
midwest |> filter(state == "wi")         # capitalization matters
midwest |> filter(state == "WI" | "MI")  # has to be state == "MI"
```

- The first one needs a double equals sign. 
  - A single `=` *sets* something, like an argument (`digits = 0`) or a new column in `mutate()`. 
  - A double `==` *asks* a question, "is this equal to that?", and answers `TRUE` or `FALSE` for every row. `filter()` needs the question.
- The last one looks like it should work, but R needs a full comparison on each side of `|`. Use `%in%` instead.

## `arrange()`: Sort Rows {.smaller}

:::: {.columns}

::: {.column width="50%"}
Default: smallest first

```{r}
midwest |>
  arrange(percbelowpoverty) |>
  select(county, state, percbelowpoverty)
```
:::

::: {.column width="50%"}
With `desc()`: largest first

```{r}
midwest |>
  arrange(desc(percbelowpoverty)) |>
  select(county, state, percbelowpoverty)
```
:::

::::

## `slice()`: Pick Rows by Position {.smaller}

:::: {.columns}

::: {.column width="50%"}
`slice()`: rows by number

```{r}
midwest |>
  slice(1:3) |>
  select(county, state, poptotal)
```
:::

::: {.column width="50%"}
`slice_sample()`: random rows

```{r}
midwest |>
  slice_sample(n = 3) |>
  select(county, state, poptotal)
```
:::

::::

- `slice_head(n = 5)` and `slice_tail(n = 5)`: the first or last 5 rows
- `slice_sample()` gives different rows every time you run it
- Position depends on order, so `arrange()` then `slice_head()` gives you a "top 5." `slice_max()` and `slice_min()` do that in one step (more on those after the break).

## `distinct()` and `count()` {.smaller}

:::: {.columns}

::: {.column width="50%"}
`distinct()`: just the unique values

```{r}
midwest |> distinct(state)
```
:::

::: {.column width="50%"}
`count()`: unique values *and* how many rows have each

```{r}
midwest |> count(state)
```
:::

::::

Add `sort = TRUE` to `count()` to put the biggest first.

## `select()`: Keep Some Columns {.smaller}

```{r}
midwest |>
  select(county, state, starts_with("perc"))
```

- Pick columns by name, or with helpers like `starts_with()`, `ends_with()`, and `contains()`
- A minus sign drops columns: `select(-PID)`
- You can also use column numbers (called indexes): `select(1, 2:5)`

## `rename()` and `relocate()` {.smaller}

```{r}
midwest |>
  rename(metro = inmetro, pop = poptotal) |>
  relocate(metro, .after = state) |>
  select(county:pop)
```

- `rename(new_name = old_name)`
- `relocate()` moves columns without dropping any
- Why the dot in `.after`? Most arguments to dplyr verbs are **your column names**, so dplyr puts a `.` in front of its own settings (`.after`, `.before`, `.by`) to keep the two from colliding. If your data had a column called `after`, then `relocate(after)` would still mean that column.

## `mutate()`: Make New Columns {.smaller}

```{r}
midwest |>
  mutate(
    pop_thousands = poptotal / 1000,
    metro_label = if_else(inmetro == 1, "Metro", "Non-metro")
  ) |>
  select(county, state, pop_thousands, metro_label)
```

## `if_else()`: Pick Between Two Values {.smaller}

```{r}
#| eval: false
if_else(inmetro == 1, "Metro", "Non-metro")
```

- `if_else()` takes 3 arguments, in this order:
  - a **condition**: a test that's `TRUE` or `FALSE` for each row
  - the value to use where the condition is `TRUE`
  - the value to use where the condition is `FALSE`
- Here, R checks each county one at a time: if `inmetro == 1` is `TRUE`, the new column gets `"Metro"`; if it's `FALSE`, it gets `"Non-metro"`
- You'll also see base R's `ifelse()` (no underscore) online. It works the same way, but `if_else()` is stricter and catches more mistakes.

## `case_when()`: More Than Two Outcomes {.smaller}

::: {.callout-tip title="For reference"}
We won't use `case_when()` again in class, but it's here for when you need it.
:::

:::: {.columns}

::: {.column width="60%"}
```{r}
#| eval: false
midwest |>
  mutate(
    size = case_when(
      poptotal >= 100000 ~ "Large",
      poptotal >= 25000 ~ "Medium",
      .default = "Small"
    )
  ) |>
  count(size)
```
:::

::: {.column width="40%"}
```{r}
#| echo: false
midwest |>
  mutate(
    size = case_when(
      poptotal >= 100000 ~ "Large",
      poptotal >= 25000 ~ "Medium",
      .default = "Small"
    )
  ) |>
  count(size)
```
:::

::::

- Each line is `condition ~ value`. Read the `~` as "then".
- R checks the lines **top to bottom** for each row, and the first `TRUE` wins. That's why `"Medium"` doesn't need `& poptotal < 100000`.
- `.default` catches everything that didn't match (another dot argument!)

## Your Turn (15 min) {.smaller}

Using `midwest`, write one pipeline for each:

1. Non-metro counties in Illinois, sorted from highest to lowest poverty rate
2. Keep only `county`, `state`, and every column that starts with `pop`
3. Add a column for the share of adults with a college degree **who are not** professionals (`percollege - percprof`)

Work with a neighbor. If you get an error, read the whole message aloud to your partner first. It often tells you exactly what's wrong.

## Your Turn: One Way to Do It {.smaller}

```{r}
#| eval: false
# 1
midwest |>
  filter(state == "IL", inmetro == 0) |>
  arrange(desc(percbelowpoverty))

# 2
midwest |>
  select(county, state, starts_with("pop"))

# 3
midwest |>
  mutate(college_nonprof = percollege - percprof)
```

## Break {background-color="#007155"}

# dplyr, Part 2: Groups {background-color="#007155"}

## `group_by()` + `summarize()` {.smaller}

```{r}
midwest |>
  group_by(state) |>
  summarize(
    counties = n(),
    total_pop = sum(poptotal),
    avg_poverty = mean(percbelowpoverty)
  )
```

- `group_by()` doesn't change what you see. It changes what happens *next*.
- `summarize()` collapses each group down to **one row**
- `count()` did this, but `summarize()` gives a lot more options!

## Grouping by Two Things {.smaller}

```{r}
midwest |>
  group_by(state, inmetro) |>
  summarize(avg_poverty = mean(percbelowpoverty), .groups = "drop")
```

**A question we might have**: is non-metro poverty higher than metro poverty in every state?

## Watch Out: Averaging Averages (or Percentages) {.smaller}

`mean(percbelowpoverty)` treats a county of 1,700 people the same as Cook County (5 million). To get the real state-wide rate, add up the people, *then* divide:

```{r}
midwest |>
  group_by(state) |>
  summarize(
    poverty_rate = sum(poptotal * percbelowpoverty / 100) / sum(poptotal) * 100
  )
```

This is where data and coding collide, and it will come up again with the NBRC data.

## Missing Values {.smaller}

- Real data has gaps. In R, a missing value is `NA`.
- Any math with an `NA` in it returns `NA`, unless you tell R to skip them

```{r}
mean(c(10, 20, NA))
mean(c(10, 20, NA), na.rm = TRUE)
```

- `midwest` has no missing values, but your BEA data might. The HW2 'starter code' handles most of it.

## Dropping Missing Values: `is.na()` {.smaller}

- To drop rows with a missing value, use `is.na()`. The `!` means "not".

```{r}
town_income <- tibble(
  town = c("Burlington", "Montpelier", "St. Albans"),
  income = c(52000, NA, 61000)
)

town_income |> filter(!is.na(income))
```

- `filter(income == NA)` **doesn't work**. Nothing is ever "equal to" a missing value, so always use `is.na()`.

## `slice_max()` and `slice_min()` {.smaller}

:::: {.columns}

::: {.column width="50%"}
The **biggest** county in each state:

```{r}
midwest |>
  group_by(state) |>
  slice_max(poptotal, n = 1) |>
  select(county, state, poptotal)
```
:::

::: {.column width="50%"}
The **smallest** county in each state:

```{r}
midwest |>
  group_by(state) |>
  slice_min(poptotal, n = 1) |>
  select(county, state, poptotal)
```
:::

::::

- `# Groups: state [5]` means the result is **still grouped**. Anything you do next happens state by state until you add `ungroup()`.
<!-- - Shortcut: many verbs can group for just one step, with nothing left over:
  `summarize(avg_poverty = mean(percbelowpoverty), .by = state)`. The `slice_` verbs spell it `by`, with no dot -->

## Saving What You Make

```{r}
rural_wi <- midwest |>
  filter(state == "WI", inmetro == 0)

nrow(midwest)
nrow(rural_wi)
```

- Without `<-`, the result prints and then disappears
- **Give results new names** rather than overwriting the original. If you write `midwest <- midwest |> filter(...)`, the rest of the data is gone until you re-run from the top.

## Your Turn (15 min)

1. How many non-metro counties does each state have? (Hint: `count()` can take two columns.)
2. For each state, find the non-metro county with the highest share of college graduates
3. **This one's challenging:** for each state, what share of *people* (not counties) live in a metro county?

## Your Turn: One Way to Do It {.smaller}

```{r}
#| eval: false
# 1
midwest |> count(state, inmetro)

# 2
midwest |>
  filter(inmetro == 0) |>
  group_by(state) |>
  slice_max(percollege, n = 1)

# 3
midwest |>
  group_by(state) |>
  summarize(metro_share = sum(poptotal[inmetro == 1]) / sum(poptotal))
```

# Code Style {background-color="#007155"}

## Why Style Matters

- You will read your code more often than you write it, and so will your NBRC teammates
- R4DS Ch. 4 is the full guide. Here are the rules that matter most.

## A Few Rules (More Like Guidelines) {.smaller}

```{r}
#| eval: false
# hard to read
df2<-midwest|>filter(state=="WI",inmetro==0)|>mutate(PopK=poptotal/1000)

# easier
rural_wi <- midwest |>
  filter(state == "WI", inmetro == 0) |>
  mutate(pop_k = poptotal / 1000)
```

- Names in `snake_case`, and names that describe the contents
- Spaces around `<-`, `==`, `+`, and after commas
- One verb per line, with `|>` at the end of each line
- Comments explain **why**, not what

## Reading Error Messages {.smaller}

| You see | It usually means |
|---|---|
| `could not find function "glimpse"` | You didn't run `library(tidyverse)` |
| `object 'midwst' not found` | A typo, or you never ran the chunk that created it |
| `unexpected symbol` | A missing comma, parenthesis, or `|>` |
| `We detected a named input` | You used `=` where you meant `==` |

Read the **whole** message. The answer is usually in there.

# HW2 Preview {background-color="#007155"}

## HW2: Wrangling Data (Due Sept. 27)

- Download real county data from the Bureau of Economic Analysis (BEA)
- Use at least three dplyr verbs from tonight
- Download a second BEA table and **join** the two (joins are next week)
- Full instructions are in your `hw2.qmd`

## Let's Do Step 1 Together

- BEA's website takes a few clicks to navigate, so we'll get everyone's `hw2a.csv` downloaded and uploaded into `homework/data` before you leave
- After that, everything else in the assignment uses tonight's tools

## Before Next Class

- [ ] **HW1** due Sunday, Sept. 13 by 11:59pm. Render before you save!
- [ ] Download your first BEA file for HW2, if we didn't finish tonight
- [ ] Read R4DS [Ch. 7 (Data import)](https://r4ds.hadley.nz/data-import) and [Ch. 8 (Getting help)](https://r4ds.hadley.nz/workflow-help), and skim [Ch. 5 (Data tidying)](https://r4ds.hadley.nz/data-tidy)

Next week: getting data from files, APIs, and the web, plus joins.

## Thank you! {background-color="#007155"}
